Agda Meets Accelerate

### 2012-07-31 paper/main.tex
section 4 and 5

@@

Accelerate relies on Haskell's type class \texttt{Num} and built-in
magic to embed constants. The magic is that Haskell scans each integer
literal as a value of type \texttt{Integer}, which is a built-in type
of unlimited integers. To this scanned value, Haskell applies the
function \texttt{fromInteger} to convert to the expected
type. Similarly, floating point constants are scanned as values of
type \texttt{Rational} (\texttt{Integer} fractions) and then converted
using \texttt{fromRational}. Accelerate provides suitable instances of
the numeric type classes that define \texttt{fromInteger} and
\texttt{fromRational} to produce suitable AST fragments.

In Agda, the situation is more involved because it only supports
natural number literals which abbreviate a unary encoding. Therefore,
we embed a numeric literal using a string with an explicit
type annotation that determines the parsing of the string. Here are
some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
It should be noted that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Element} as all other Haskell
types, which can be referred to in an Accelerate program:
\begin{verbatim}
data Element : Set where
  Bool : Element
  Int : Element
  Float : Element
  Double : Element
  ...
\end{verbatim}
All magic of the embedding is hidden in the \texttt{:::} operation:
\begin{verbatim}
_:::_ : (s : String) -> (E : Element) 
     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The argument \texttt{nu} is an instance argument that is automatically
filled-in with a suitably typed value in scope. Here, the predicate
\texttt{IsNumeric} plays the role of a type class that characterizes
the numeric types.
\begin{alltt}
IsNumeric : Element -> Set
IsNumeric Int = \(\top\)
IsNumeric Float = \(\top\)
IsNumeric Double = \(\top\)
IsNumeric _ = \(\bot\)
\end{alltt}
The function \texttt{parsesAs} dispatches on its ``type'' argument and
checks whether the string is a constant of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries, thus
they are computed from \texttt{E} using the function \texttt{EltDict}
and \texttt{ReadDict}.

@@

In a number of places, Accelerate's generativity limits the
applicability of dependent typing. We already mentioned that the
formalization of associativity or of the concept monoid gets
unmanageable because it has to be asserted for abstract syntax.

For a similar reason, an implementation of the \texttt{filter}
operation creates problems. The problem is that the size of the result
cannot be determined statically.

Existential types do not help with this problem.

However, an alternative encoding dependently-typed of arrays can be
used which is compatible with filtering of elements. Actually, there
are several ways to implement such an encoding with different drawbacks.

>>>>

\subsection{Embedding of Haskell Types}
\label{sec:embedding-constants}

Accelerate supports a wide range of numeric types as base types for
array computations. Almost all of these types have no suitable
counterpart in Agda, which only supports natural numbers in unary
encoding. For that reason, our embedding keeps these types 
abstract in Agda. To specify functions that are polymorphic in such a
Haskell type or depend on it in some way, we have reified these types
in an Agda type \texttt{Element}.
\begin{verbatim}
data Element : Set where
  Bool   : Element
  Int    : Element
  Float  : Element
  Double : Element
  Pair   : Element -> Element -> Element
  ...
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric type is defined by a predicate \texttt{IsNumeric}.
\begin{alltt}
IsNumeric : Element -> Set
IsNumeric Int = \(\top\)
IsNumeric Float = \(\top\)
IsNumeric Double = \(\top\)
IsNumeric _ = \(\bot\)
\end{alltt}

@@

Accelerate relies on Haskell's built-in support for the type classes
\texttt{Num} and \texttt{Fractional} to embed constants. The Haskell compiler reads each integer
literal as a value of type \texttt{Integer}, which is a built-in type
of arbitrary precision integers. To this value, Haskell applies the
function \texttt{fromInteger} that converts to the expected
type. Similarly, floating point constants are read as values of
type \texttt{Rational} (\texttt{Integer} fractions) and then converted
using \texttt{fromRational}. Accelerate provides instances of
these type classes that define \texttt{fromInteger} and
\texttt{fromRational} to produce suitable AST fragments.

Because of Agda's limited support for numeric data types,
we embed more ambitious numeric literals for floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Element}.
All magic of the embedding is hidden in the \texttt{:::} operation:
\begin{verbatim}
_:::_ : (s : String) -> (E : Element) 
     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are to be inferred by Agda.
The argument \texttt{nu} is an instance argument that is automatically
filled-in with a suitably typed value in scope
\cite{DevriesePiessens2011}. Here, the predicate 
\texttt{IsNumeric} plays the role of a type class that characterizes
the numeric types.

The function \texttt{parsesAs} dispatches on its ``type'' argument and
checks whether the string is a constant of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries,
which are computed from \texttt{E} using the functions \texttt{EltDict}
and \texttt{ReadDict}.

@@

In a number of places, Accelerate's generativity limits the
applicability of dependent typing. We already mentioned that the
formalization of associativity or of the concept monoid gets
unmanageable because such properties have to be asserted for abstract syntax.

For a similar problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : {n m : Nat}{E : Element}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result
cannot be determined statically. In fact, the only thing we know about
\texttt{m} is that it must be less than or equal to
\texttt{n}. However, we cannot prove this from the code because the
Accelerate implementation computes the length of the result in a quite
subtle way. Moreover, as Accelerate deals with the construction of
abstract syntax most of the time, the value of \texttt{m} is simply
not available at run time of \texttt{filter}.

One might contemplate that an existential type like
\begin{verbatim}
exists (\ m : Nat -> m <= n -> Vector m E)
\end{verbatim}
could be employed. However, it is not possible to build such an
existential package because the evidence \texttt{m} is not available
when the package is constructed.

However, an alternative encoding of arrays can be
used which is compatible with filtering of elements. The idea of this
encoding is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to wrap each element in a maybe type.
pair up each element with a boolean flag that indicates
its presence.\footnote{A \texttt{Maybe} type would be a better option,
  but it has to be coded without using pattern matching.}
\begin{verbatim}
FVector : Nat -> Element -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}

Now filtering becomes quite simple.
\begin{verbatim}
filterF : {n : Nat}{E : Element}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
However, mapping becomes more complicated because it either has to
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too.\footnote{This
  complication could be avoided with the \texttt{Maybe} type.}
\begin{verbatim}
mapF : {n : Nat}{E F : Element}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                     else (pair (fst bx) defaultF)
\end{verbatim}
On the positive side, some operations can get rid of the absent
elements. In particular, a fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : {n : Nat}{E : Element}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Other operations like \texttt{fold1} and the scan operations present
in Accelerate can also be lifted to this representation, but they
retain a notion of absent elements and do not allow to revert to a
non-filtered representation.

In the end, such a representation may not be a loss on a GPU. As long
as all computations take the same path, all processing elements work
in unison. As soon as there are different paths in the same
computation step, then some elements will be idle for part of the
computation step. So it would be most advantageous to organize work as
uniformly as possible.
